---
 output: github_document
---
 
 <!-- README.md is generated from README.Rmd. Please edit that file -->
 
```{r, include = FALSE}
 knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   fig.path = "man/figures/README-",
   out.width = "100%"
 )
```
 
 # saasi
 
 <!-- badges: start -->
 <!-- badges: end -->
 
Saasi is an ancestral state reconstruction method that accounts for variation in sampling rates among locations or traits. 
 
 ## Installation
 
 You can install the development version of saasi from
 [GitHub](https://github.com/) with:
 
 ``` r
 # install.packages("remotes")
 remotes::install_github("MAGPIE-SFU/saasi")
 ```
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(saasi)
library(diversitree)
library(tidytree)
library(ape)
library(phytools)
library(readr)

```

## saasi package

This is a demo showing how to use the saasi package.

saasi requires the following argument:
(1) a phylogenetic tree (class `phylo`). The tree should satisfy the following 
properties: rooted, binary, branch length in units of time and positive, 
has tree$tip.state and has no missing states. Check `check_tree_compatibility` function
for details. In addition, `prepare_tree_for_saasi` function helps the user to 
prepare a tree that is compatible with saasi.
(2) speciation, extinction and sampling rates (class `data.frame`). These rates 
can be estimated using the function `estimate_bds_parameters`.
(3) a transition rate matrix (class `matrix`). 
The output will be a data frame that contains the probabilities 
of each state for each internal node of the phylogenetic tree. The Q matrix 
can be estimated using the function `estimate_transition_rates`.


## Example: Ebola 2013-2016 West African Ebola Epidemic


For the next example, download the data from Nextstrain:
https://nextstrain.org/ebola/ebov-2013?c=country


```{r Nextstrain Ebola}
# Read the tree file and metadata (replace with your own directory)
tree <- read.tree("testing_data/nextstrain_ebola_ebov-2013_timetree.nwk")
metadata <- read_tsv("testing_data/nextstrain_ebola_ebov-2013_metadata.tsv")

# Create a data.frame contains strains and states. 
# The first column should match tree$tip.label, 
# the second column is the region or country that you are interested in.
tip_data <- data.frame(
  tip_label = metadata$strain,
  state = metadata$country
)

# Check compatibility
check_tree_compatibility(tree)

# The tree is not compatible. Specifically, the tree is nonbinary, it does
# not contains tip states, contains negative/zero branch length, and contains
# polytomy.
# To fix it, run the following command:

ebola_tree <- prepare_tree_for_saasi(tree,tip_data)

# Check compatibility again
check_tree_compatibility(ebola_tree)

# The tree is compatible with saasi.

# Estimate the transition rate matrix Q
Q <- estimate_transition_rates(ebola_tree,method = 'simmap',model = 'SYM')

# Estimate the rates
rates <- estimate_bds_parameters(
    ebola_tree,
    mu = 5,
    r0_max = 3,
    r0_min = 1.5,
    psi_max = 15,
    infectious_period_min = 20/365, # convert to years
    infectious_period_max = 40/365, # convert to years
    n_starts = 100)

# Setting up parameters
pars1 <- create_params_template(colnames(Q),lambda = rates$lambda,mu = rates$mu,psi = rates$psi)

# Run saasi analysis
saasi_ebola <- saasi(ebola_tree,pars1,Q)

# Plot and save the result
p1 <- plot_saasi(ebola_tree,saasi_ebola,save_file = "ebola_equal_psix.png")

# Different set up, we assume Liberia samples less than other countries
pars2 <- create_params_template(colnames(Q),lambda = rates$lambda,mu = rates$mu,psi = c(rates$psi,rates$psi/2,rates$psi))

# Rerun analysis and save the result
saasi_ebola2 <- saasi(ebola_tree,pars2,Q)

p2 <- plot_saasi(ebola_tree,saasi_ebola2,save_file = "ebola_unequal_psix.png")

```

```{r Sanity check}
# Sanity check, Q gets large if we randomly shuffle the tip.state

testing_ebola <- ebola_tree
set.seed(123) 
testing_ebola$tip.state <- sample(testing_ebola$tip.state)

Q <- estimate_transition_rates(tesing_ebola,method = 'simmap',model = 'SYM')
# qij gets extremely large

```
