---
 output: github_document
---
 
 <!-- README.md is generated from README.Rmd. Please edit that file -->
 
```{r, include = FALSE}
 knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   fig.path = "man/figures/README-",
   out.width = "100%"
 )
```
 
 # saasi
 
 <!-- badges: start -->
 <!-- badges: end -->
 
Saasi is an ancestral state reconstruction method that accounts for variation in sampling rates among locations or traits. 
 
 ## Installation
 
 You can install the development version of saasi from
 [GitHub](https://github.com/) with:
 
 ``` r
 # install.packages("remotes")
 remotes::install_github("MAGPIE-SFU/saasi")
 ```
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(saasi)
library(diversitree)
library(tidytree)
library(ggtree)
library(ggplot2)
library(ggimage)
```

## saasi package

This is a demo showing how to use the saasi package.

To run saasi, we need (1) a phylogenetic tree (class `phylo`), (2) speciation, 
extinction and sampling rates (class `data.frame`), and (3) a transition rate matrix
(class `matrix`). The output will be a data frame that containing the probabilities 
of each state for each internal node of the phylogenetic tree.

## A simulation
We will first simulate a tree with known rates, and known internal node states, to illustrate saasi's ancestral state inferences. 

We simulate a birth-death-sampling tree, for which we need to specify speciation, 
extinction, sampling rates and transition rates.

```{r parameters}
pars <- data.frame(state=c(1,2),prior=c(0.5,0.5),lambda=c(3,3),mu=c(0.05,0.05),psi=c(.1,1))

qij_matrix <- function(k) {
  mat <- matrix(0.15, nrow = k, ncol = k)  
  diag(mat) <- NA  
  return(mat)
}
q_matrix = qij_matrix(2)

set.seed(1)

phy <- sim_bds_tree(pars, q_matrix, x0=1, max_taxa = 300, max_t = 300,
             include_extinct = FALSE)

plot(phy)

```

## Modifying tree

You might notice that the tree also include all the tips at the present day.
This is because the simulation stopped at the maximum allowed time. In pathogen phylogenetics and phylogeography applications, we typically do not have heterochronous sequences (from the present). In this simulation, we therefore drop the tips at the present day. 


```{r modified}

k=2
phy <- prune(phy)
h <- history.from.sim.discrete(phy, 1:k)

true_phy_info <- as_tibble(phy)
phy_data <- c(factor(h$tip.state),factor(h$node.state))
true_phy_info$State <- phy_data
true_phy <- ggtree(phy)
true_phy <- true_phy  %<+% true_phy_info + geom_point(aes(color=State),size=2) +
  ggtitle("True Phylogeny") +
  theme(text = element_text(size = 15,family = "serif"),plot.title = element_text(size=15))
true_phy
node_depths <- node.depth.edgelength(phy)
tmrca <- max(node_depths)
tips_to_drop <- phy$tip.label[abs(node_depths[1:length(phy$tip.label)] - tmrca) <= 0.01]
new_phy <- drop.tip(phy, tips_to_drop)
phy_our <- new_phy
true_phy_info_new <- as_tibble(new_phy)
phy_data <- c(factor(h$tip.state),factor(h$node.state))
true_phy_info_new <- true_phy_info_new %>% mutate(State = phy_data[label])
new_phy$tip.state <- new_phy$tip.state[setdiff(names(new_phy$tip.state), tips_to_drop)]
phy <- new_phy
true_phy_new <- ggtree(phy)
true_phy_new <- true_phy_new  %<+% true_phy_info_new + geom_point(aes(color=State),size=2) +
  ggtitle("True Phylogeny - without present day tips") +
  theme(text = element_text(size = 15,family = "serif"),plot.title = element_text(size=15))
true_phy_new
```

## Ancestral state inference

Now we can use the simulated tree to do ancestral state inference.
The function `ace` in the `ape` package does ancestral character (here, state) estimation without considering sampling rates, and it is a natural comparison for saasi since it is widely used in large-scale phylogeographic reconstructions. Comparing `saasi`'s reconstructions to `ace`'s illustrates the impact of adjusting for sampling differences. 



```{r ace}
library(ape)
ace_phy <- phy
ace_phy$node.label <- NULL
# Note: Do not have this problem if use earlier version `ape`
# Error in names(obj$ace) <- phy$node.label : 
# attempt to set an attribute on NULL

ace_phy$tip.state <- ace_phy$tip.state[setdiff(names(ace_phy$tip.state), tips_to_drop)]
asr_ace<-ace(ace_phy$tip.state, ace_phy,type = "discrete", model="ER")

ace_node_lik <- as.data.frame(asr_ace$lik.anc)
ace_node_lik$node <- 1:new_phy$Nnode + Ntip(new_phy)

ace_pie <- nodepie(ace_node_lik,cols=1:k)

p1 <- ggtree(ace_phy)
p1 <- p1 %<+% true_phy_info_new + geom_tippoint(aes(color=State),size=2)+
  ggtitle("ace") +
  theme(text = element_text(size = 15,family = "serif"),plot.title = element_text(size=15))
p1 <- inset(p1, ace_pie,width = 0.07,height = 0.07,hjust=0.005)
p1



```


We see that `ace` would infer that most of the internal nodes are in State 2 
instead of State 1.

Now let's try `saasi`.



```{r saasi}
result <- saasi(phy,pars,q_matrix)
node_result <-  result[-(1:22), ]
node_result
node_result$node <- 1:ace_phy$Nnode + Ntip(ace_phy)
our_pie <- nodepie(node_result,cols=1:k)

p2 <- ggtree(ace_phy)
p2 <- p2 %<+% true_phy_info_new + geom_tippoint(aes(color=State),size=2)+
  ggtitle("SAASI") +
  theme(text = element_text(size = 15,family = "serif"),plot.title = element_text(size=15))
p2 <- inset(p2, our_pie,width = 0.07,height = 0.07,hjust=0.005)
p2
```
Due to accounting for the sampling differences `saasi` infers most of the internal nodes correctly. 

## Parameter estimations - Speciation and Extinction

Suppose we do not know the speciation and extinction rates for each state, but 
we have some knowledge about the sampling rates (e.g. per year/month, this should
algin with your tree time). We estimate the 
speciation and extinction rates using the method described in
\href{https://doi.org/10.1093/molbev/msr217}{Stadler et al. (2012)}.

```{r estimates}

estimates <- mle_lm(phy,lambda = 2, mu = 0.1, psi = 1,lower = c(0.001,0.001), upper = c(5,5))

estimates[1]
estimates[2]
```


This MLE approach should be robust to different initial guesses, for example:

```{r estimates alternative}

estimates <- mle_lm(phy,lambda = 100, mu = 100, psi = 1,lower = c(0.001,0.001), upper = c(500,500))

estimates[1]
estimates[2]
```
However, sometimes the algorithm might cause an error: Error: L-BFGS-B needs
finite values of `fn' of Complex Objective Function. This is due to a large 
branch length (the value exceed the .Machine$double.xmax), we need to set the upper bound smaller. 

TODO: for future versions, will find a more consistent way of finding the 
estimates.

## Parameter estimations - Transition

If the transition rates are also unknown, one easy way of estimating transition rates
is using `ace`:

```{r estimates qij}

# a function that convert the ace estimates to a matrix (one of the inputs in saasi)
replace_matrix_with_vector <- function(matrix, vector) {
  for (i in 1:nrow(matrix)) {
    for (j in 1:ncol(matrix)) {
      matrix[i,j] <- vector[matrix[i,j]]
    }
  }
  return(matrix)
}

# TODO: will add this to the utility function

q_matrix_est <- replace_matrix_with_vector(asr_ace$index.matrix,asr_ace$rates)
q_matrix_est
```

Now we rerun `saasi` with estimated parameters.

```{r rerun saasi with estimated parameters}
pars_est <- data.frame(state=c(1,2),
                   prior=c(0.5,0.5),
                   lambda=rep(estimates[1],2),
                   mu=rep(estimates[2],2),
                   psi=c(.1,1))

result <- saasi(phy,pars_est,q_matrix_est)

# now draw the plot

node_result <-  result[-(1:22), ]
node_result
node_result$node <- 1:ace_phy$Nnode + Ntip(ace_phy)
our_pie <- nodepie(node_result,cols=1:k)

p3 <- ggtree(ace_phy)
p3 <- p3 %<+% true_phy_info_new + geom_tippoint(aes(color=State),size=2)+
  ggtitle("SAASI - estimated parameters") +
  theme(text = element_text(size = 15,family = "serif"),plot.title = element_text(size=15))
p3 <- inset(p3, our_pie,width = 0.07,height = 0.07,hjust=0.005)
p3

```
The result is slightly different than using the true parameters.