---
title: "Sampling Aware Ancestral State Inference"
author: "Yexuan Song"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Introduction to SAASI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
  <!-- setwd("/Users/yexuan-magpie/Desktop/MSc_thesis_implementation/saasi") -->
---
 
 <!-- README.md is generated from README.Rmd. Please edit that file -->
 
```{r, include = FALSE}
 knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   fig.path = "man/figures/README-",
   out.width = "100%"
 )
```
 
# Introduction
 
<!-- badges: start -->
<!-- badges: end -->
 

This vignette serves as a guide of how to use the `saasi` package.

Note: In the future version it will also include a guide of how to 
visualize the inference results (phylogenetic tree  and 
alluvial plot visualization).

<!-- This vignette serves two major proposes: -->

<!-- (1). How to do the ancestral state inference using -->
<!-- sassi on a phylogenetic tree of interest. -->

<!-- (2). How to visualize the inference results on the tree. -->


# saasi

## Phylogenetic tree 

The tree that you are interested in to infer ancestral states using `saasi'
should be a timed - rooted binary tree, including tip states
(either geographic locations or some certain traits). The traits should be 
discrete traits.

In this vignette, we will go through saasi's workflow using 
a bird tree example in the `ape` package.

```{r bird order data}
library(ape)
library(saasi)
data(bird.orders)

# set some random discrete characters to the bird tree.
x <- factor(c(rep(1, 5), rep(2, 18)))

bird.orders$tip.state <- x
```

## Getting input parameters

In addition to a phylogenetic tree, saasi also requires the user to know the 
phylodynamic parameters (speciation, extinction, transition & sampling).

Note that the user should have some knowledge about the sampling difference 
between states of interest (or for testing different sampling scenarios).

The speciation and extinction rates (the time unit should be consistent 
with the time scale of the tree) can be estimated using "mle_lm" function. 
The method is described in 
[Stadler et al. (2012)](https://doi.org/10.1093/molbev/msr217).


```{r speciation and extinction}
estimates <- mle_lm(bird.orders,lambda = .2, mu = .05, psi = .1,lower = c(0.001,0.001), upper = c(0.5,0.5))

estimates
```


Transition rates can be estimated using `ace'.

```{r tranistion}
ace_qij <- extract_ace_q(ace(x, bird.orders, type = "d"))
```

## Running saasi

Now we can run `saasi` function. 

```{r saasi}
# set up the parameters
pars <- data.frame(state=c(1,2),prior=c(0.5,0.5),lambda=c(.08,.08),mu=c(0.001,0.001),
                   psi=c(.01,.09))

saasi_result <- saasi(bird.orders,pars,ace_qij)
```


## transition rate adjustment

Sometimes the transition rate adjustment using `ace` is biased due to uneven
sampling. We implement a quick way of adjusting the transition rates based on
the sampling assumptions.

Suppose in this example State 1 samples 1/9.

```{r transition adj}

adj_qij <- q_adjust(ace_qij,"1",1/9)
saasi_result_with_adjusted_qij <- saasi(bird.orders,pars,adj_qij)

```


# Visualization

A quick way of checking the result is by using the default functions 
(`tiplabels` & `nodelabels`) in `ape`.

```{r visualization}
plot(bird.orders, label.offset = 1)

color <- c("#E41A1C", "#377EB8")

# Map tip states to colors
tip_colors <- color[bird.orders$tip.state]
tiplabels(bg = tip_colors, cex = 1, adj = 1, pch = 21)

nodelabels(pie = saasi_result, 
           piecol = color[1:ncol(saasi_result)], 
           cex = 0.2)# plotting

```

```{r visualization2}
plot(bird.orders, label.offset = 1)

color <- c("#E41A1C", "#377EB8")

# Map tip states to colors
tip_colors <- color[bird.orders$tip.state]
tiplabels(bg = tip_colors, cex = 1, adj = 1, pch = 21)

nodelabels(pie = saasi_result_with_adjusted_qij, 
           piecol = color[1:ncol(saasi_result_with_adjusted_qij)], 
           cex = 0.2)# plotting

```

<!-- There are two major ways of visualizing saasi's output: -->

<!-- (1). Phylogenetic tree, internal nodes are pie charts represent probabilities. -->

<!-- (2). Alluvial plot, showing the overall movements between states. -->

<!-- ## Phylogenetic tree visualization -->

<!-- ### Tree visualization using ggtree -->

<!-- ### Tree visualization using `ape` build-in functions  -->


<!-- ## Alluvial plot visualization -->


